---
title: "Palmer Penguin Data cleaning"
author: "Michael Olagoke"
date: "2023-08-21"
output:
  html_notebook: default
  pdf_document: default
  html_document: default
---

## Introduction 
  The `palmerpenguin` is a package in R and like all other packages in R, it comes with test dataset which are `penguins` and `penguins_raw`. In this project, I leveraged on my knowledge of R to clean up the `penguins` before further analysis. Load the `palmerpenguins` package to access the `penguins` data.
```{r}
# install the package if you do not have it on your local machine

# load the package
library(palmerpenguins)

# Quick peep at the data
View(penguins)
```
  
  ### About the `Penguins` data
  The data is a collection of 344 records of penguins observed over a period of 3 years(2007 - 2009). Each include species of penguins, Islands they reside, flipper length in mm, bill length in mm, bill depth in mm, sex and year the observation was recorded.
```{r}
# load the data into a variable so that the source of data is preserved
data <- penguins
# summary of the penguins data
summary(data)

# View the first 6 values
head(data)
 
```
  ## Data Cleaning
  From the result above, it is clear that `null` (na) values in some columns, alternatively i built my `custom count function` to count na values by columns.
```{r}
# custom custom function - count_na
count_na <- function (a){
  counts <- sapply(a,function (x) sum(is.na(x)))
  data.frame(counts)
}

# run the function with 'data' as parameter
count_na(data)
```
  
  #### Replacing na within quantitative column with mean of the total column.
  Each column has consistent value such as a fixed decimal place for quantitative column.I made a function that calculated the mean and formats the result to maintain consistency. Then I assigned nulls in each column with the mean of the respective column.
```{r}
# mean function. It takes the column and desired decimal place as parameters
cal_mean <- function (a,b){
  as.double(round(mean(a, na.rm=TRUE),b))
}

# replacing nulls with mean

# bill_depth_mm
data$bill_depth_mm[is.na(data$bill_depth_mm)] <- cal_mean(data$bill_depth_mm,1)
# bill_length_mm
data$bill_length_mm[is.na(data$bill_length_mm)] <- cal_mean(data$bill_length_mm,1)
# flipper_length_mm
data$flipper_length_mm[is.na(data$flipper_length_mm)] <- cal_mean(data$flipper_length_mm,0)
data$body_mass_g[is.na(data$body_mass_g)] <- cal_mean(data$body_mass_g,0)

```
  Run the count_na function to check if na values still exist
```{r}
count_na(data)

# summarize the data to check for NA's
summary(data)
```
  Data has been cleaned!.
  